---
output: github_document
---

# Most Popular Unisex Names

```{r setup}
library(tidyverse)
library(gt)
library(gtExtras)
```

## Unisex Names

There are some names that are used commonly by both the sexes. Let us dive into 
the data and figure out what are the most popular unisex names. Let us 
assume that a name is considered unisex if there are more than 33% males and 
33% females with that name. We can tweak these thresholds subsequently to see 
if it reveals a different set of names!

### Import

```{r import-data}
FILE_NAMES <- here::here('data/names.csv.gz')
tbl_names <- readr::read_csv(FILE_NAMES, show_col_types = FALSE)
tbl_names
```

### Transform

```{r tbl-names-unisex-1}
tbl_names_unisex_1 <- tbl_names |> 
  group_by(name, sex) |> 
  summarize(
    nb_births = sum(nb_births),
    .groups = 'drop'
  )

tbl_names_unisex_1
```

```{r tbl-names-unisex-2}
tbl_names_unisex_2 <- tbl_names_unisex_1 |> 
  pivot_wider(
    names_from = sex, 
    names_prefix = 'nb_births_',
    values_from = nb_births, 
    values_fill = 0
  )

tbl_names_unisex_2
```

```{r tbl-names-unisex-3}
tbl_names_unisex_3 <- tbl_names_unisex_2 |> 
 mutate(
   nb_births = nb_births_M + nb_births_F,
   pct_births_M = nb_births_M / nb_births,
   pct_births_F = nb_births_F / nb_births
 )

tbl_names_unisex_3
```


```{r tbl-names-unisex-4}
tbl_names_unisex_4 <- tbl_names_unisex_3 |> 
  filter(
    pct_births_M > 0.33,
    pct_births_F > 0.33,
    nb_births > 50000
  ) |> 
  arrange(desc(nb_births))

tbl_names_unisex_4
```

```{r tbl-names-unisex-5}
tbl_names_unisex_5 <- tbl_names_unisex_4 |> 
  select(name, nb_births, pct_births_M, pct_births_F) |> 
  group_by(name, nb_births) |> 
  nest() |> 
  ungroup() |>
  head(15)

tbl_names_unisex_5
```


```{r tbl-names-unisex}
tbl_names_unisex <- tbl_names |> 
  # Aggregate births across `name` and `sex`
  group_by(name, sex) |> 
  summarize(
    nb_births = sum(nb_births),
    .groups = 'drop'
  ) |>
  # Pivot the table wider
  pivot_wider(
    names_from = sex, 
    names_prefix = 'nb_births_',
    values_from = nb_births, 
    values_fill = 0
  ) |> 
  # Add columns for total births and percentage male and female
  mutate(
    nb_births = nb_births_M + nb_births_F,
    pct_births_M = nb_births_M / nb_births,
    pct_births_F = nb_births_F / nb_births
  ) |> 
  # Filter for popular unisex names 
  # [At least 33% males and females, > 50000 total births]
  filter(
    pct_births_M > 0.33,
    pct_births_F > 0.33,
    nb_births > 50000
  ) |> 
  arrange(desc(nb_births)) |> 
  select(name, nb_births, pct_births_M, pct_births_F) |> 
  group_by(name, nb_births) |> 
  nest() |> 
  ungroup() |>
  head(15)

tbl_names_unisex
```

```{r}
tbl_names_unisex |>
  gt::gt() |> 
  gtExtras::gt_plt_bar_stack(
    data, 
    width = 65,
    labels = c("MALE", "FEMALE"),
    palette= c("#2596be", "#f4ba19"),
    fmt_fn = scales::label_percent()
  ) |> 
  fmt_number(columns = 'nb_births', decimals = 0) |> 
  gt::tab_header(
    title = md("**The Most Common Unisex Names**"),
    subtitle = "Names for which at least one-third the names were male, and
    at least one-third were female, through 2021"
  ) |> 
  gtExtras::gt_theme_538()
```


### Transform


```{r}
tbl_names_unisex <- tbl_names |> 
  # Group by name and sex
  group_by(name, sex) |> 
  # Summarize the total number of births, drop the grouping
  summarize(nb_births = sum(nb_births), .groups = 'drop') |> 
  # Group by name
  group_by(name) |> 
  # For each name, add NEW columns with number and pct of births
  mutate(
    nb_births_total = sum(nb_births),
    pct_births = nb_births / nb_births_total
  ) |> 
  # Remove the grouping
  ungroup() |> 
  # Keep rows corresponding to Males
  filter(sex == "M") |> 
  # Keep only names with more than 50,000 births and pct between 0.33 and 0.67
  filter(
    between(pct_births, 0.33, 0.67),
    nb_births_total > 50000
  )

tbl_names_unisex
```

### Visualize

```{r}
tbl_names_unisex_trends <- tbl_names |> 
  semi_join(tbl_names_unisex, by = "name") |> 
  group_by(name, year) |> 
  mutate(
    nb_births_total = sum(nb_births),
    pct_births = nb_births / nb_births_total
  ) |> 
  filter(sex == "M") |> 
  group_by(name) |> 
  summarize(
    nb_births_total = sum(nb_births_total),
    pct_births_by_year = list(pct_births),
    pct_births_M = sum(nb_births) / sum(nb_births_total),
    pct_births_F = 1 - pct_births_M,
    .groups = 'drop'
  ) |> 
  group_by(name, nb_births_total, pct_births_by_year) |> 
  nest(pct_births = c(pct_births_M, pct_births_F)) |> 
  ungroup() |> 
  arrange(desc(nb_births_total)) |> 
  select(name, nb_births_total, pct_births, pct_births_by_year)
  
tbl_names_unisex
```

```{r}
tbl_names_unisex_trends |> 
  gt::gt() |> 
  fmt_number(nb_births_total, decimals = 0) |> 
  gtExtras::gt_plt_bar_stack(
    pct_births,
    width = 65,
    labels = c("MALE", "FEMALE"),
    palette= c("#2596be", "#f4ba19"),
    fmt_fn = scales::label_percent()
  ) |> 
  gtExtras::gt_plt_sparkline(pct_births_by_year) |> 
  gt::tab_header(
    title = md("**The Most Common Unisex Names**"),
    subtitle = "Names for which at least one-third the names were male, and
    at least one-third were female, through 2021"
  ) |> 
  gtExtras::gt_theme_538()
```

