---
output: github_document
---
## Combining Data

__Combining__ data spread across multiple tables is a common task. This is where the `join` and `bind` functions in `dplyr` comes into play. 


```{r mindmap-combining-data, echo = FALSE, out.width = "100%"}
knitr::include_graphics(
  path = "https://i.imgur.com/qfpUdFQ.png"
)
```



```{r setup, message = FALSE}
# Load packages and common utility functions
library(tidyverse)
source(here::here("_common.R"))
```

### Join

__Joining__ tables means combining the rows of one table with the rows of another based on common columns. In `dplyr`, you can use the `inner_join()`, `left_join()`, `right_join()`, and `full_join()` functions to perform different types of joins. 

Let us take two datasets `band_instruments` and `band_members` which contain information on the instruments that different members play and the bands that they belong to. Note that not everyone plays an instrument, and not everyone is in a band.

```{r data-band-instruments}
band_instruments
```

```{r data-band-members}
band_members
```

#### `inner_join()`: Keep only matching rows in both tables

The `inner_join()` function returns only the rows that have matching values in both tables. In this case, we get a table with only those members who play an instrument and belong to a band.

```{r inner-join}
inner_join(band_instruments, band_members, by = "name")
```


#### `left_join()`: Keep all rows in the left table

The `left_join()` function returns all the rows from the left table and the matching rows from the right table. Hence, we have all members who play an instrument, even if they did not belong to a band.

```{r left-join}
left_join(band_instruments, band_members, by = "name")
```


#### `right_join()`: Keep all rows in the right table

The `right_join()` function returns all the rows from the right table and the matching rows from the left table. Hence, we have all members who belong to a band, even if they did not play an instrument.

```{r right-join}
right_join(band_instruments, band_members, by = "name")
```


#### `full_join()`: Keep all rows in both tables

The `full_join()` function returns all rows from both tables, filling in `NA` for missing values. Here, we get all members whether or not they play an instrument or belong to a band.

```{r full-join}
full_join(band_instruments, band_members, by = "name")
```

#### `cross_join()`: Cartesian join of both tables

The `cross_join()` function is a cartesian join that matches every row in one table with every row in another table WITHOUT consideration for any matching columns. As a result, we end up with a table with number of rows equaling the product of number of rows of the two tables.

```{r cross-join}
cross_join(band_instruments, band_members)
```

#### `semi_join()`: Keep columns with matching rows in the right table

The `semi_join()` function returns all rows from the left table with a match in the right table. It is very similar to the `inner_join()`, but differs in that it ONLY returns the columns in the left table.

```{r semi-join}
semi_join(band_instruments, band_members, by = "name")
inner_join(band_instruments, band_members, by = "name")
```

#### `anti_join()`: Keep columns with non-matching rows in the right table

The `anti_join()` function returns all rows from the left table that DONT have a match in the right table. You can think of it as the opposite of `semi_join()`.


```{r anti-join}
anti_join(band_instruments, band_members, by = "name")
```
### Bind

While the `join` functions combine two tables by matching rows, the `bind` functions combine them by stacking them along rows or columns.

#### `bind_rows()`: Bind multiple data frames by row

```{r bind-rows}
df1 <- tibble(x = 1:2, y = letters[1:2])
df2 <- tibble(x = 4:5, z = 1:2)
bind_rows(df1, df2)
```


#### `bind_cols()`: Bind multiple data frames by column

```{r bind-cols}
df1 <- tibble(x = 1:3)
df2 <- tibble(y = 3:1)
bind_cols(df1, df2)
```

