---
output: github_document
---


```{r setup, include = FALSE, message = FALSE}
library(tidyverse)
knitr::opts_chunk$set(comment = "#>")
```

## Aggregating Data

Let us read the `babynames` data frame first so we can use it to explore data aggregation.

```{r read-babynames}
NAMES_FILE <- here::here("data/names.csv.gz")
babynames <- readr::read_csv(NAMES_FILE, show_col_types = FALSE)
babynames
```

### `group_by`: Group by one or more variables

The `group_by()` function lets you group a table by one or more columns. Applying the `group_by` function to a table does not change its appearance, except for adding some metadata on the grouping variables.

```{r}
# Group by ONE column
babynames |> 
  group_by(sex)
```

```{r}
# Group by multiple columns
babynames |> 
  group_by(sex, name)
```


### `summarize`: Summarize each group down to one row

The reason we group a table by columns is to be able to manipulate it by group. While the `group_by` function can be used in tandem with several other manipulation functions, the most common function used is the `summarize` function. It aggregates the data for each group such that there is only ONE row per group.

You can use any arbitrary function inside `summarize` as long as it collapses multiple values into a single value. For example, we can summarize the `babynames` dataset for each `sex` in terms of the total number of births and number of distinct names.

```{r}
babynames |> 
  group_by(sex) |> 
  summarize(
    nb_births = sum(nb_births),
    nb_names = n_distinct(name)
  )
```

How about we try grouping by multiple variables? Let us compute the same summaries, but grouping by `sex` and `year` this time.

```{r}
babynames |> 
  group_by(sex, year) |> 
  summarize(
    nb_births = sum(nb_births),
    nb_names = n_distinct(name)
  )
```

It is very important to pay attention to the message that gets printed.

```
#> `summarise()` has grouped output by 'sex'. You can override using the `.groups`
#> argument.
```

When you group by multiple variables, each summarize operation removes only the last level of grouping. For example, in this case, the grouping by `year` was removed, but the grouping by `sex` is still retained. This design has some advantages as it lets us do hierarchical summaries without having to invoke multiple `group_by` operations. However, it can also introduce nasty surprises if you are not careful. 

My recommendation would be to override this default behavior and always drop all grouping after a summarize operation. This might add some overhead especially if you are going to group it again. But doing that explicitly will save you a lot of pain.

```{r}
babynames |> 
  group_by(sex, year) |> 
  summarize(
    nb_births = sum(nb_births),
    nb_names = n_distinct(name),
    .groups = "drop"
  )
```


