---
output: github_document
---
## Aggregating Data

__Aggregating__ data involves summarizing the data at a higher level than the original data, such as computing the `mean`, `sum`, or `count` of values for different `groups` of data. The two primary functions that you will use to aggregate data are:

```{r mindmap-aggregate-data, echo = FALSE, out.width = "100%"}
knitr::include_graphics(
  path = "https://i.imgur.com/BMjwi2P.png"
)
```


- `group_by()`: This function is used to group data by one or more variables. The result is a `grouped_df` object, which can then be used with other `dplyr` functions to aggregate the data.

- `summarize()` or `summarise()`: These functions are used to compute summary statistics for each group of data. For example, you can compute the `mean()`, `median()`, `sum()`, or `count()` of values in each group.

Additionally, the grammar of data manipulation provided by `tidyverse` allows us to utilize the full power of data manipulation operations like `mutate()`, `filter()`, `arrange()` etc. to grouped data as well.


Let us load the `tidyverse` and read the babynames data so we can use it to explore data aggregation.


```{r setup, message = FALSE}
# Load packages and common utility functions
library(tidyverse)
source(here::here("_common.R"))
```

```{r read-babynames}
file_name_names <- here::here("data/names.csv.gz")
tbl_names <- readr::read_csv(file_name_names, show_col_types = FALSE)
tbl_names
```

### `group_by()`: Group by one or more variables

The `group_by()` function lets you group a table by one or more columns. Applying the `group_by` function to a table does not change its appearance, except for adding some metadata on the grouping variables.

```{r group-by-one-column}
# Group by ONE column
tbl_names |> 
  group_by(sex)
```

```{r group-by-multiple-columns}
# Group by multiple columns
tbl_names |> 
  group_by(sex, name)
```


### `summarize()`: Summarize each group down to one row

The reason we group a table by columns is to be able to manipulate it by group. While the `group_by` function can be used in tandem with several other manipulation functions, the most common function used is the `summarize` function. It aggregates the data for each group such that there is only ONE row per group.

You can use any arbitrary function inside `summarize` as long as it collapses multiple values into a single value. For example, we can summarize the `babynames` dataset for each `sex` in terms of the total number of births and number of distinct names.

```{r summarize-by-sex}
tbl_names |> 
  group_by(sex) |> 
  summarize(
    nb_births = sum(nb_births),
    nb_names = n_distinct(name)
  )
```

How about we try grouping by multiple variables? Let us compute the same summaries, but grouping by `sex` and `year` this time.

```{r summarize-by-sex-year}
tbl_names |> 
  group_by(sex, year) |> 
  summarize(
    nb_births = sum(nb_births),
    nb_names = n_distinct(name)
  )
```

It is very important to pay attention to the message that gets printed.

```
#> `summarize()` has grouped output by 'sex'. You can override using the `.groups`
#> argument.
```

When you group by multiple variables, each summarize operation removes only the last level of grouping. For example, in this case, the grouping by `year` was removed, but the grouping by `sex` is still retained. This design has some advantages as it lets us do hierarchical summaries without having to invoke multiple `group_by` operations. However, it can also introduce nasty surprises if you are not careful. 

My recommendation would be to override this default behavior and always drop all grouping after a summarize operation. This might add some overhead especially if you are going to group it again. But doing that explicitly will save you a lot of pain.

```{r summarize-by-sex-year-drop}
tbl_names |> 
  group_by(sex, year) |> 
  summarize(
    nb_births = sum(nb_births),
    nb_names = n_distinct(name),
    .groups = "drop"
  )
```

While, you can apply any of the data manipulation verbs in the `tidyverse` to a grouped data frame, there are some that are going to be used more than the others.

### `mutate()`: Create, modify, and delete columns

A grouped data frame is just like a normal data frame EXCEPT that any operation you do on it is applied to each group separately. For example, suppose you want to compute the percentage of births in a year by `sex` and `name`. You can group by `sex` and `name` and use the `mutate()` function to add a column for the total number of births for each group, and then use that to compute the percentages. 

Let us now `ungroup()` the data, `filter()` for names with more than 50,000 births and `arrange()` them in descending order of births, so we can see how some names have a lion's share of births happening in a few years. For example, 40% of all Lunas were born in the years 2019, 2020, and 2021


```{r mutate-nb-births}
tbl_names |> 
  group_by(sex, name) |> 
  mutate(
    nb_births_total = sum(nb_births),
    pct_births_in_year = nb_births / nb_births_total
  ) |> 
  ungroup() |> 
  filter(nb_births_total > 50000) |> 
  arrange(desc(pct_births_in_year))
```

Note that it is very important to ungroup the data using `ungroup()` before applying any data frame wide operations. It is more efficient to operate on whole data frames and grouped operations are always going to be slower.

### `across()`: Apply a function across multiple columns

Finally, there are situations where you might want to apply a function across multiple columns. This is where the `across()` function comes very handy. You can use different selector functions available to summarize a group of columns without having to write multiple lines of code. Shown below are two examples, where we summarize the mean of a number of columns of the data by `cut`.

- In the first example, we select the columns to summarize across by name
- In the second example, we select the columns to summarize across by their data type.

```{r across}
# Summarize multiple columns by name
diamonds |> 
  group_by(cut) |> 
  summarize(across(c(carat, price), mean))

# Summarize multiple columns by data type
diamonds |> 
  group_by(cut) |> 
  summarize(across(where(is.numeric), mean))
```

